---
title: "DESeq2 Fulvestrant"
output:
  html_document: default
  always_allow_html: true
  word_document: default
---

Update: 20th November 2020

# Differential gene expression using DESeq2

Differential gene expression using DESeq2 and visualisation with Venn diagrams. Using Adam's code, some from https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('functions.R')
library(org.Hs.eg.db)
library(AnnotationDbi)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(DT)
library(limma)
library(pheatmap)
library(readr)
# to convert to HTML document, type in Console > rmarkdown::render(input = "file_name.Rmd", output_format = "html_document", output_file = NULL) - have to do HTML due to tab layout
```


```{r}
# need to edit metadata sheet to merge condition 1 and 2 together
metadata_fulv <- read.table("metadata_fulv.csv", sep=",", header = TRUE)
rownames(metadata_fulv) <- metadata_fulv$Samples
colnames(txi.fulv$counts) <- rownames(metadata_fulv)
all(rownames(metadata_fulv) %in% colnames(txi.fulv$counts))
```


```{r}
# only need to separate metadata if comparing different patients

# set up meta_data
# meta_data_1 <- meta_data %>% 
  # filter(Cell_line == "MCF") %>% 
  # column_to_rownames("Samples")

```



```{r, message=FALSE, eval=FALSE}
## Set up the DESeq2 object
df_mRNA <-  txi.fulv
deseq2_txi <- function(df_mRNA, metadata_fulv, control="3_Control", test="3_100nM_fulv", value){
    dds<- DESeqDataSetFromTximport(df_mRNA, metadata_fulv, ~Condition2)
  
  keep <- rowSums(counts(dds)) >= 10
  dds <- dds[keep,]
  
  dds <- DESeq(dds)
  
  res <- results(dds, contrast = c(value, test, control))
  
  return(res)
}
res_day3 <- deseq2_txi(df_mRNA, metadata_fulv, control = "3_Control", test="3_100nM_Fulv", value="Condition2")

res_day7 <- deseq2_txi(df_mRNA, metadata_fulv, control = "7_Control", test="7_100nM_Fulv", value="Condition2")

res_fulv <- deseq2_txi(df_mRNA, metadata_fulv, control = "3_100nM_Fulv", test="7_100nM_Fulv", value="Condition2")
```

The following describes the analysis of the data using DEseq2. I have separated the analyses into three different contrasts and have plotted summary statistics and MA plots for each.  


## Summary of the data

### Day 3 Control v Fulvestrant

```{r}
summary(res_day3, alpha=0.05)
```

### Day7 Control v Fulvestrant

```{r}
summary(res_day7, alpha=0.05)
```

### Day 3 Fulvestrant v Day 7 Fulvestrant

```{r}
summary(res_fulv, alpha=0.05)
```


## MA plots 

In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be coloured blue if the adjusted p value is less than 0.01. Points which fall out of the window are plotted as open triangles pointing either up or down.

### Day 3 Control v Fulvestrant

```{r, fig.width=8, fig.height=5}
DESeq2::plotMA(res_day3)
```

### Day 7 Control v Fulvestrant

```{r, fig.width=8, fig.height=5}
DESeq2::plotMA(res_day7)
```

### Day 3 v Day 7

```{r, fig.width=8, fig.height=5}
DESeq2::plotMA(res_fulv)
```






## Results tables

The folowing results tables show the significant genes (p<0.05). Filtering has been performed with a log2 fold change +/- 2. These have been exported as .csv files where _res.csv is all genes, and _sig.csv is significant genes (p<0.05) with LFC +/-2 (see results folder)


### Day 3 Control v Fulvestrant
```{r, message=FALSE}
dt <- filter_genes(as.data.frame(res_day3), name="fulv_day3_control_v_fulv")
datatable(dt$sig)
```


### Day 7 Control v Fulvestrant

```{r, message=FALSE}
dt <- filter_genes(as.data.frame(res_day7), name="fulv_day7_control_v_fulv")
datatable(dt$sig)
```

### Day 3 Fulvestrant v Day 7 Fulvestrant

```{r, message=FALSE}
dt <- filter_genes(as.data.frame(res_fulv), name="fulv_day3fulv_v_day7fulv")
datatable(dt$sig)
```
