---
title: "Gene Ontology Analysis - Untreated"
output:
  word_document: default
  pdf_document: default
  html_document: default
---


# Gene ontology and gene set enrichment analysis
Pathway analysis using code from Rbio Session 4. Gene ontology using cluster Profiler https://yulab-smu.github.io/clusterProfiler-book/

Update: 5th November 2020


```{r, include=FALSE}
library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
#Go to Rbio.Rmd file to load full list, including installing packages etc. If having trouble, install individual packages using CRAN within R Studio. If still having trouble, then have to search Bioconductor for it.
```
## Day 0 v Day 3

### Filter data (process data)
The data contains double entries or NA entries. We need to remove these for downstream analysis.

```{r}
library(readr)
day0v3_sig <- read_csv("results/day0v3_sig.csv")

# Use Ensembl IDs to map Entrez Gene IDs. Use #keytypes(org.Hs.eg.db) to check what column names are
day0v3_sig$Entrez.Gene <- mapIds(org.Hs.eg.db, keys=as.character(day0v3_sig$Row.names), keytype="ENSEMBL", column="ENTREZID")
#Keep only rows from table without NAs
day0v3_sig <- day0v3_sig[is.na(day0v3_sig$Entrez.Gene)==FALSE,]
#Remove duplicated entries
day0v3_sig <- day0v3_sig[!duplicated(day0v3_sig$Entrez.Gene),]
```

### Gene Ontology (GO) over-representation test
We are going to perform an over-representation test for our genes in GO terms from org.Hs.eg.db

```{r}
#Change Entrez IDs from numbers to characters
geneset <- as.character(day0v3_sig$Entrez.Gene)
#This will show Entrez IDs only
head(geneset)
#This will take a little while to run
day0v3_ego <- enrichGO(gene = geneset, 
                universe = NULL, #all available genes in database
                OrgDb = org.Hs.eg.db, #Hs: homo sapiens
                ont ="BP", #molecular function, biological process, cellular component
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,  #q value is FDR adjusted p value
                readable = TRUE) #will show gene symbol in images later rather than Entrez Gene ID
#dimensions - number of GO terms
dim(day0v3_ego)

day0v3_GOgenes <- data.frame(day0v3_ego$ID, day0v3_ego$Description, day0v3_ego$p.adjust, day0v3_ego$geneID)
#Copy and paste into Excel
```
### Showing the ego object in graphical form

```{r,fig.height=4, fig.width=6}
barplot (day0v3_ego, showCategory = 10)
dotplot (day0v3_ego, showCategory=10)
#can use simplify() to remove redundant terms, but can be slow eg barplot(simplify(ego))
```

We can look into the relation of the enriched GO terms using the `emapplot()` function. You will need to increase the height and width of the image so that the terms don't overlap.  

```{r, fig.height=8, fig.width=10}
emapplot(day0v3_ego)
```

We can use cnetplot() to display the relation of specific genes within the GO terms.

```{r, fig.height=12, fig.width=14}
#logFC is optional, but if available will colour-code the nodes
logFC_mcf <- day0v3_sig$log2FoldChange
#set name of object
names(logFC_mcf) <- day0v3_sig$Entrez.Gene
cnetplot(day0v3_ego, foldChange = logFC_mcf, showCategory = 5, circular = FALSE)
#show 5 GO terms, if set circular to TRUE then GO terms won't be in middle, will be on edge of circle
#size = number of genes, fold change = red for upregulated, green for downregulated
```
